Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys

### 2019-01-24 doc/article.tex
forging on with edits

In the next decade, SNe~Ia will be used as peculiar-velocity probes to measure the the influence of gravity on structure formation within the Universe.
The peculiar velocity power spectrum is sensitive to the growth of structure as $P_{vv}\propto (fD)^2$, where $D$ is  the linear growth factor and
$f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate  \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}. 
The $\Lambda$CDM prediction for the $z=0$ velocity power spectrum is shown in Figure~\ref{power:fig}. The growth of structure depends on gravity, indeed 
\citet{2007APh....28..481L} find that General Relativity, $f(R)$,  and DGP gravities 
give
$f \approx \Omega_M^\gamma$ with $\gamma=0.55, 0.42, 0.68$ respectively.  Adopting this parameterization for gravity, peculiar velocity
surveys are  sensitive to $fD=\Omega_M^{\gamma} \exp{\left(\int_a^1 \Omega_M^{\gamma} d\ln{a} \right)}$ \citep[see Figure~2 of][to see
$fD$ plotted as a function of redshift for $\Lambda$CDM]{1475-7516-2013-04-031}.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{src/new2.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Volume-weighted velocity power spectrum $k^3P_{vv}$ at $z=0$ for $\mu=1, 0,5$ as predicted for General Relativity in the linear regime.
The bottom solid horizontal lines show the approximate range of $k$ expected to be used in surveys with corresponding
redshift depths $z_{max}$.
\label{power:fig}}
\end{figure}

Peculiar velocity surveys have already been  used to measure $fD$, though not at a precision to distinguish gravity models.
 \citet{2017MNRAS.471..839A} use 6dFGS peculiar velocities using the Fundamental Plane method and get distances to
get a 15\% uncertainty in $fD$ at $z\approx 0$ using density and velocity cross-correlations.
The upcoming 
TAIPAN survey will obtain Fundamental Plane galaxies with densities of $n_g \sim 10^{-3}h^3$\,Mpc$^{-3}$,
and the WALLABY+WNSHS surveys will obtain Tully-Fisher galaxies at densities $n_g \sim 2\times 10^{-2} - 10^{-4} h^3$\,Mpc$^{-3}$ from
$z=0-0.1$.
Covering 75\% of the sky to a maximum depth of $z=0.1$, these surveys combined are projected to have 3\% $fD$ uncertainties \citep{2017MNRAS.464.2517H}.
Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\citep{2008MNRAS.389L..47A,2014MNRAS.444.3926J,2015JCAP...12..033H, 2017JCAP...05..015H}.
For reference, DESI projects 10\% precision of $fD$ at $z>0.3$ redshifts  using RSD alone.
%Measurement of the velocity field using LSST-discovered
%SNe~Ia has been quantified by \citet{2011PhRvD..83d3004B,2017JCAP...01..060O}

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands, over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
with number density  $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$
 for which good light curves can be measured. This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.

For example, over the course of a decade an LSST SN survey can produce interesting $fD$ measurements in several
redshift bins.   
\citet{2017ApJ...847..128H} find 4--14\% uncertainties in $fD$ in $0.05$ redshift bins from $z=0$ to 0.3, where at
$0<z<0.2$ most of the signal comes from peculiar velocities and at higher redshifts $RSD$ is more constraining.

Our scientific interest is in probing gravity, so while most of the literature cast peculiar velocity results in terms of 
$fD$, the interesting parameter is $\gamma$.  Survey design should be tailored toward minimizing the uncertainty
in 
$\sigma_\gamma$ rather than $\sigma_{\ln{fD}}$.  To illustrate the distinction,
 $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega + \int \Omega^{\gamma} \ln{\Omega} d\ln{a}$, which for $\Lambda$CDM
 $\approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ respectively; given two surveys with the same percentage precision  in $fD$, the one at lower redshift
better constrains $\gamma$.  We thus thus relate surveys to $\sigma_\gamma$, where we assume a constant $\gamma$
over the redshift range of the survey.

%
%The combination $fD$ is sensitive to gravity. The growth rate $f$ describes how fast overdensities increase, which
%depends on how fast gravity makes mass move.
%General Relativity predicts
%$f \approx \Omega_M^\gamma$ with $\gamma=0.55$, whereas other gravity models can be similarly described
%with different values of
%$\gamma=0.42, 0.68$ for $f(R)$,  and DGP gravities respectively 
%\citep{2007APh....28..481L}.

The uncertainties in $\gamma$ are estimated using a
Fisher matrix analysis
similar  to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}.
The analysis is performed in Fourier space, 
where the Fisher information matrix of parameters $\lambda \in \{\gamma,bD\}$ with
the  tracer density and velocity power spectra used as data is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}.
\label{cov:eq}
\end{equation}
The parameter dependence comes in through $P_{vv}\propto ((fD)(\gamma)\mu)^2$, the SN~Ia power spectrum $P_{\delta \delta }\propto ((bD) + (fD)(\gamma)\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  ((bD) + (fD)\mu^2)f\sigma_8$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.

The uncertainty in $\gamma$ is $\sqrt{\left(F^{-1}\right)_{\gamma,\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or deviations
from $f=\Omega_M^{\gamma}$, such observations provide additional leverage in probing gravity
but are not considered here.

The survey parameters that enter the calculation are the solid angle $\Omega$, depth of the survey $r_{max}$ (or $z_{max}$), the intrinsic magnitude dispersion 
through $\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$, and the survey duration $t$ through $n=\epsilon rt$ where $r$ is the observer SN~Ia rate
and $\epsilon$ is the sample-selection efficiency.
An analysis in which the RSD and velocity surveys are treated independently (e.g.\ they do not cover the same volume) is considered by
setting the off-diagonal elements of $C$ to zero.

We consider fiducial peculiar velocity surveys for a range of redshift depths $z_{max}$, with survey durations or $t=2$ and 10~years, and with or without
cross-correlations accounted for.
The other survey parameters
$\Omega=3$, $\epsilon=0.65$, $\sigma_M=0.08$~mag are fixed.
A TAIPAN-like
galaxy survey with $b=1.2$ and number density $n_g=10^{-3}h^3$\,Mpc$^{-3}$ is used for the RSD measurement.
The uncertainties in $\gamma$ are shown in Figure~\ref{var:fig}, which shows a number of interesting features:
\begin{itemize}
\item Even shallow, two-year surveys give gammas for 
\end{itemize}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{src/var.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{
\label{var:fig}}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{src/dvardz.png}
\includegraphics[width=0.47\textwidth]{src/dvardlnt.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{
\label{dvar:fig}}
\end{figure}

@@

The $I/I_{max}$ of these  scenarios are shown in the bottom two panels of Figure~\ref{power:fig}.
Generally, $k<0.01$ modes have $I/I_{max}\sim 1$ and so are sample-noise limited.
Individual $k>0.01$ modes have  $I/I_{max}<1$ meaning  that shot noise
is significant.  Recall that collectively, $k>0.01$ modes are the important contributors to the peculiar-velocity signal.
Basically all  supernovae found in  the next decade contribute to improving the statistical measurement of $f\sigma_8$.
%The figure also shows how the statistical gain from additional supernovae diminish as the survey duration increases, and when they are at lower-redshift  (by virtue of their lower intrinsic velocity dispersion).

{\bf Solid Angle $\Omega$:}
$F_{f\sigma_{8}\,f\sigma_{8}}$ is  proportional to volume and hence the survey solid angle $\Omega$.  The baseline
LSST survey covers 18,000 sq.~deg, at $-75 \lesssim \delta \lesssim15$ avoiding the Galactic plane.  Larger solid-angle coverage beyond
the LSST baseline benefits peculiar-velocity science.  Complementary northern-hemisphere surveys and LSST-expanded or independent
coverage of the southern equatorial pole could significantly increase
the sky coverage with a corresponding decrease in the variance of $f\sigma_8$.

%s
% and so helps illustrate 
%how how the density, $n$, and intrinsic SN magnitude dispersion of peculiar velocity probes, as
%$\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$, affect survey precision.

{\bf Common-Volume Galaxy  Survey:}
Galaxy and velocity surveys in different volumes do not have the off-diagonal cross-correlation terms in Eq.~\ref{cov:eq}.
The degradation in probative power   $I/I_{max}$, when sample-variance limited,
is shown as the black dashed curves in the bottom two panels of Figure~\ref{power:fig}.
The curves lie between $0.1-0.2$, showing
that accounting for the common dark-matter fluctuations that give rise to galaxy densities and peculiar velocities
reduces the variance in $f\sigma_8$ by factors of 5 to 10.

{\bf Redshift Depth $z_{max}$:}
Increasing the survey redshift depth increases the $f \sigma_8$ precision, but when resource-limited is the expense of additional depth worth neglecting lower redshift
SNe? The answer is found in  Table~\ref{tab:howlett}, which shows that after 10 years of LSST the lowest redshifts have the
strongest probative power and are sensitive more to peculiar velocities rather than galaxy overdensities.  Despite
the differential volume increase with redshift depth (Eq,~\ref{fisher:eqn}), the degradation in intrinsic peculiar velocity $\propto z/(1+z)$
and being shot-noise-limited  makes a low-redshift tracer more valuable than one at higher redshift.

{\bf Intrinsic Magnitude Dispersion $\sigma_M$}:
Even after 10-years, a SN peculiar velocity survey still has significant shot noise.
Thus, a decrease in the supernova intrinsic magnitude dispersion leads to a significant improvement in  the STON of $f\sigma_8$.
%as shown in Figure~\ref{notshot:fig} in parallel to our earlier discussion of the dependence of $n$.  We are not sample-variance
%limited so the $\sigma^2_M/n$-term
%in the Fisher matrix is important.
Survey follow-up strategy,
which determines which data are available per SN~Ia, influences the accuracy to which their distances can be determined. 
This dependence is non-trivial: the improvement between $\sigma_M=0.08$ and 0.15~mag dispersions is equivalent to a factor of 3.52
 in number density, or equivalently in survey duration.

%\subsection{Survey Parameters that Determine Scientific Performance}

@@

%
%
%\begin{align}
%C^{-1}_{,s}  = &  \frac{1}{(P_{gg}(k,\mu) + \frac{1}{n_g})(P_{vv}(k,\mu) + s)- P_{vg}(k,\mu)^2}
% \begin{bmatrix}
%   1 &
%   0\\
%   0 &
%  0
%   \end{bmatrix}.\\
%  & -
% \frac{ P_{gg}(k,\mu) + \frac{1}{n_g} }{\left((P_{gg}(k,\mu) + \frac{1}{n_g})(P_{vv}(k,\mu) + s)- P_{vg}(k,\mu)^2\right)^2} \\
%& \times 
%     \begin{bmatrix}
%     P_{vv}(k,\mu) + s &
%   -P_{vg}(k,\mu)  \\
%  - P_{vg}(k,\mu)  &
%  P_{gg}(k,\mu) + \frac{1}{n_g} 
%   \end{bmatrix}.
% \end{align}

>>>>

In the next decade, SNe~Ia will be used as peculiar-velocity probes to measure  the influence of gravity on structure formation within the Universe.
The peculiar velocity power spectrum is sensitive to the growth of structure as $P_{vv}\propto (fD)^2$, where $D$ is  the linear growth factor and
$f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate  \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}.\footnote{
To be precise, the peculiar
velocity power spectrum has dependence $P_{vv}\propto (HfD)^2$.  A supernova  survey  measures luminosity distance fluctuations
$\delta_{d_L} = (d_L- \bar{d}_L(z))/\bar{d}_L(z)$, where $d_L$ is the observed distance and $\bar{d}_L(z)$ is the expected distance at the observed redshift $z$.
To first order in peculiar velocity along the line of sight $v$, $\delta_{d_L} =  v \left(1-\frac{1}{H\bar{d}_L(z)}\right) \approx -\frac{v}{H\bar{d}_L(z)}$
at low redshift.   The $H$-dependence of  $P_{vv}$ and the conversion from distances to velocities cancel, making  peculiar velocity surveys sensitive to $(fD)^2$.
}
The $\Lambda$CDM prediction for the $z=0$ velocity power spectrum is shown in Figure~\ref{power:fig}. The growth of structure depends on gravity, indeed 
\citet{2007APh....28..481L} find that General Relativity, $f(R)$,  and DGP gravities 
follow the relation
$f \approx \Omega_M^\gamma$ with $\gamma=0.55, 0.42, 0.68$ respectively.  Adopting this parameterization to model gravity, peculiar velocity
surveys are  sensitive to $fD=\Omega_M^{\gamma} \exp{\left(\int_a^1 \Omega_M^{\gamma} d\ln{a} \right)}$, whose $\Lambda$CDM prediction is plotted 
in Figure~2 of  \citet{1475-7516-2013-04-031}.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{src/new2.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Volume-weighted velocity power spectrum $k^3P_{vv}(z=0)$ for $\mu=1, 0.5$ (solid black, cyan)  as predicted for General Relativity in the linear regime.
Volume-weighted peculiar velocity shot-noise  $k^3\sigma^2/n$ at $z=0.1$, expected from 2- and 10-year (red, black) supernova densities, 0.08 and 0.15~mag (dashed, dotted) intrinsic magnitude dispersions.
The bottom solid horizontal lines show the approximate range of $k$ expected to be used in surveys with corresponding
redshift depths $z_{max}$.
\label{power:fig}}
\end{figure}

Peculiar velocity surveys have already been  used to measure $fD$, though not to a level where  gravity models can be precisely distinguished.
 \citet{2017MNRAS.471..839A} use 6dFGS peculiar velocities using the Fundamental Plane method to
get a 15\% uncertainty in $fD$ at $z\approx 0$ using density and velocity cross-correlations.
The upcoming 
TAIPAN survey will obtain Fundamental Plane galaxies with densities of $n_g \sim 10^{-3}h^3$\,Mpc$^{-3}$,
and the WALLABY+WNSHS surveys will obtain Tully-Fisher galaxies at densities $n_g \sim 2\times 10^{-2} - 10^{-4} h^3$\,Mpc$^{-3}$ from
$z=0-0.1$.
Covering 75\% of the sky to a maximum depth of $z=0.1$, these surveys combined are projected to have 3\% $fD$ uncertainties \citep{2017MNRAS.464.2517H}.
(For reference, DESI projects 10\% precision of $fD$ at $z>0.3$ redshifts  using RSD alone.)
Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\citep{2008MNRAS.389L..47A,2014MNRAS.444.3926J,2015JCAP...12..033H, 2017JCAP...05..015H}.
%Measurement of the velocity field using LSST-discovered
%SNe~Ia has been quantified by \citet{2011PhRvD..83d3004B,2017JCAP...01..060O}

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands, over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
 for which good light curves can be measured, corresponding to a  number density of $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$
 (the total rest-frame volumetric SN~Ia rate being $7.84\times 10^{-5} h^3$\,Mpc$^{-3}$\,yr$^{-1}$.)
  This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.

Given these two advances, over the course of a decade an LSST SN survey can produce interesting $fD$ measurements in several
redshift bins.   
\citet{2017ApJ...847..128H} find 4--14\% uncertainties in $fD$ in $0.05$ redshift bins from $z=0$ to 0.3, where at
$0<z<0.2$ most of the signal comes from peculiar velocities and at higher redshifts $RSD$ is more constraining.

Our scientific interest is in probing gravity, so while most of the literature cast peculiar velocity results in terms of 
$fD$, we now focus on $\gamma$.  Survey design should be tailored toward minimizing the uncertainty in $\gamma$
$\sigma_\gamma$ rather than $\sigma_{\ln{fD}}$.  To illustrate the distinction,
 $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega + \int \Omega^{\gamma} \ln{\Omega} d\ln{a}$, which for $\Lambda$CDM
 $\approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ respectively; given two surveys with the same percentage precision  in $fD$, the one at lower redshift
better constrains $\gamma$. 
The uncertainties in $\gamma$ are estimated using a
Fisher matrix analysis
similar  to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}.
The analysis is performed in Fourier space, 
where the Fisher information matrix of parameters $\lambda \in \{\gamma,bD\}$ is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}.
\label{cov:eq}
\end{equation}
The parameter dependence enters through $fD$ in the relations $P_{vv}\propto (fD\mu)^2$, the SN~Ia overdensity
power spectrum $P_{\delta \delta }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.  Note that while the $bD$ term does contain information on $\gamma$, we do not use it in this analysis.
The uncertainty in $\gamma$ is $\sigma_\gamma = \sqrt{\left(F^{-1}\right)_{\gamma,\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or non-constant $\gamma$, such observations provide additional leverage in probing gravity
but are not considered here.

The survey parameters that enter the calculation are the solid angle $\Omega$, depth of the survey $r_{max}$ (or $z_{max}$), the intrinsic SN~Ia magnitude dispersion 
through $\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$, and the survey duration $t$ through $n=\epsilon rt$ where $r$ is the observer SN~Ia rate
and $\epsilon$ is the sample-selection efficiency.
An analysis in which the RSD and velocity surveys are treated independently is considered by
setting the off-diagonal elements of $C$ to zero.

We calculate  $\sigma_\gamma$ of  peculiar velocity surveys for a range of redshift depths $z_{max}$, with survey durations of $t=2$ and 10~years, covering $3\pi$  of sky, and with 
analyses accounting for or neglecting
cross-correlations.
The other survey parameters
$\Omega=3$, $\epsilon=0.65$, $\sigma_M=0.08$~mag are fixed.
The bias of SN~Ia hosts is taken to be $b=1.2$.
The uncertainties in $\gamma$ are shown in Figure~\ref{var:fig}.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{src/var.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Projected $\sigma_\gamma$ of  peculiar velocity surveys for a range of redshift depths $z_{max}$, with survey durations of $t=2$ and 10~years, and with 
analyses accounting for or neglecting
cross-correlations.
\label{var:fig}}
\end{figure}

Shallow, short-duration peculiar-velocity surveys already achieve meaningful constraints on gravity.   A 2-year, $z_{max}=0.1$ survey has $\sigma_\gamma \sim 0.038$,
which can distinguish  between  General Relativity, $f(R)$,  and DGP gravities at the $2\sigma$ level.
Using cross-correlation information decreases $\sigma_\gamma$ by  factors of $\sim 0.8$.   
 This modest improvement 
comes from the suppression of  sample variance.
It is the density measurement that is sample-variance
limited,  the peculiar velocity measurement is shot-noise limited.

Having established that peculiar velocity surveys can produce a precise test of gravity, we now consider survey optimization through the 
survey parameters  $\Omega$, $z_{max}$, $t$, and $\sigma_M$, and also show the relative importance of different
redshifts with a given survey.  An important survey parameter that is not considered is cost.
\begin{description}
\item[Solid Angle $\Omega$]
The Fisher Matrix $F$ is  proportional to the survey solid angle $\Omega$.  The baseline
LSST survey covers 18,000 sq.~deg, at $-75 \lesssim \delta \lesssim15$ avoiding the Galactic plane.  Larger solid-angle coverage beyond
the LSST baseline benefits peculiar-velocity science.  Complementary northern-hemisphere surveys and LSST-expanded or independent
coverage of the southern equatorial pole could significantly increase
the sky coverage with a corresponding decrease in the variance of $\gamma$.
\item[Differential Redshift Bin $z$] Certain differential redshift bins  constrain $\gamma$ more strongly than others.  
The left panel
of Figure~\ref{dvar:fig} shows $\partial \sigma_\gamma / \partial z$  for two- and ten-year surveys with limiting $z_{max}=0.2$.
In both cases the shape of the curve is the same, with the amplitude monotonically decreasing from $z=0.01$ out to $z=0.2$.  If we had
to sculpt the distribution (say due to limited follow-up resources), the preference would be to sculpt out the highest redshift bins
resulting in a decreased $z_{max}$.   The optimal redshift distribution is thus the SN-discovery distribution and the sample is unambiguously
parameterized by $z_{max}$.  Note that this result is based on constraining a constant redshift-independent $\gamma$.
\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{src/dvardz.png}
\includegraphics[width=0.47\textwidth]{src/dvardlnt.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Left: $\partial \sigma_\gamma / \partial z$  for two- and ten-year surveys with limiting $z_{max}=0.2$.  Right: $\partial \sigma_\gamma / \partial \ln{t}$ as a function of
 $z_{max}$ for two- and ten-year surveys.
\label{dvar:fig}}
\end{figure}
\item[Redshift Depth $z_{max}$]
Increasing the survey redshift depth increases the $\gamma$ precision though that precision asymptotes
$z_{max} \sim 0.2$ as seen in Figure~\ref{var:fig}.
 \item[Survey duration $t$] There is a significant decrease on $\sigma_\gamma$ when going from a 2- to ten-year survey, as shown
 in  Figure~\ref{var:fig}.   The right panel of Figure~\ref{dvar:fig} shows $\partial \sigma_\gamma / \partial \ln{t}$ as a function of
 $z_{max}$ for two- and ten-year surveys.  The non-zero amplitudes show that shot-noise affects both two- and ten-year surveys, but
 much less so for the latter more so at low redshifts.  After 10-years, the effect of sample-variance important for shallower surveys.
\item[Intrinsic Magnitude Dispersion $\sigma_M$]
The relationship between intrinsic magnitude dispersion and precision in $\gamma$ can be seen through the 
previous result, as $\partial \sigma_\gamma/\partial \sigma_M = \partial \sigma_\gamma / \partial \ln{t}$.
Even after 10-years, a SN peculiar velocity survey still has significant shot noise.
Thus, a decrease in the supernova intrinsic magnitude dispersion leads to a significant improvement in  the STON of $f\sigma_8$.
%as shown in Figure~\ref{notshot:fig} in parallel to our earlier discussion of the dependence of $n$.  We are not sample-variance
%limited so the $\sigma^2_M/n$-term
%in the Fisher matrix is important.
Survey follow-up strategy,
which determines which data are available per SN~Ia, influences the accuracy to which their distances can be determined. 
This dependence is non-trivial: the improvement between $\sigma_M=0.08$ and 0.15~mag dispersions is equivalent to a factor of 3.52
 in number density, or equivalently in survey duration. 
 \end{description}

Shot-noise limit touched upon several times in the preceding discussion.  Go back to Figure.

There are a number of interesting features to note:
\begin{itemize}
\item Ten-year survey has lower  $\sigma_\gamma$  relative to  two-year surveys with factors that are redshift-dependent.  For $z_{max}=0.3$
the uncertainties are reduced by a factor $\sim 0.5$ whereas at  $z_{max}=0.7$ the factor is $\sim 0.7$
\end{itemize}

 Shot noise
is important  Figure~\ref{power:fig}.

@@

The $I/I_{max}$ of these  scenarios are shown in the bottom two panels of Figure~\ref{power:fig}.
Generally, $k<0.01$ modes have $I/I_{max}\sim 1$ and so are sample-noise limited.
Individual $k>0.01$ modes have  $I/I_{max}<1$ meaning  that shot noise
is significant.  Recall that collectively, $k>0.01$ modes are the important contributors to the peculiar-velocity signal.
Basically all  supernovae found in  the next decade contribute to improving the statistical measurement of $f\sigma_8$.
%The figure also shows how the statistical gain from additional supernovae diminish as the survey duration increases, and when they are at lower-redshift  (by virtue of their lower intrinsic velocity dispersion).
%\subsection{Survey Parameters that Determine Scientific Performance}

@@

\begin{align}
C^{-1}_{,\sigma_M}  =   & \frac{\ln{10}}{5} \frac{z}{1+z} \frac{2\sigma}{n}
( \frac{1}{(P_{gg}(k,\mu) + \frac{1}{n})(P_{vv}(k,\mu) + \frac{\sigma^2}{n})- P_{vg}(k,\mu)^2}
 \begin{bmatrix}
   1 &
   0\\
   0 &
  0
   \end{bmatrix}\\
   & - 
 \frac{ (P_{gg}(k,\mu) + \frac{1}{n}) }
 {\left((P_{gg}(k,\mu) + \frac{1}{n})(P_{vv}(k,\mu) + \frac{\sigma^2}{n})- P_{vg}(k,\mu)^2\right)^2} \\
& \times 
     \begin{bmatrix}
     P_{vv}(k,\mu) + \frac{\sigma^2}{n} &
   -P_{vg}(k,\mu)  \\
  - P_{vg}(k,\mu)  &
  P_{gg}(k,\mu) + \frac{1}{n} 
   \end{bmatrix})\\
 \end{align}
%
%
%\begin{align}
%C^{-1}_{,s}  = &  \frac{1}{(P_{gg}(k,\mu) + \frac{1}{n_g})(P_{vv}(k,\mu) + s)- P_{vg}(k,\mu)^2}
% \begin{bmatrix}
%   1 &
%   0\\
%   0 &
%  0
%   \end{bmatrix}.\\
%  & -
% \frac{ P_{gg}(k,\mu) + \frac{1}{n_g} }{\left((P_{gg}(k,\mu) + \frac{1}{n_g})(P_{vv}(k,\mu) + s)- P_{vg}(k,\mu)^2\right)^2} \\
%& \times 
%     \begin{bmatrix}
%     P_{vv}(k,\mu) + s &
%   -P_{vg}(k,\mu)  \\
%  - P_{vg}(k,\mu)  &
%  P_{gg}(k,\mu) + \frac{1}{n_g} 
%   \end{bmatrix}.
% \end{align}
